---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document: default
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# For reading zipped csv files
library(readr,warn.conflicts=F, quietly=T)
library(dplyr,warn.conflicts=F, quietly=T)
library(ggplot2,warn.conflicts=F, quietly=T)
library(lubridate,warn.conflicts=F, quietly=T)

# Displaying data frames using a nice format
library(lemon)
knit_print.data.frame <- lemon_print
```


## Loading and preprocessing the data

To load the data I will use the function *read_csv()* from the *readr* library to read a zipped csv file.

```{r load data}
# Read  zipped CSV into r
data <- read_csv("activity.zip", col_names = TRUE)

dim(data)
summary(data)
```


## What is mean total number of steps taken per day?

To calculate the total, mean, and median steps per day I will use the *summarise* function from *dplyr* and graph the histogram with *ggplot*.

```{r summarised_data, warning=FALSE}
# Get summary statistics by day
basic_stats_by_day <- summarise(group_by(data, date), total_steps=sum(steps), mean_steps=mean(steps, na.rm = TRUE), median_steps=median(steps,na.rm = TRUE))

# Use geom_bar to draw the histogram because data is already in frequency format
ggplot(basic_stats_by_day, aes(x=date,y=total_steps)) + geom_bar(stat = 'identity') +
  ggtitle("Total number of steps per day")
```

```{r print tables summary, render=lemon_print}
# Print mean and median steps per day, head and tail to save space
head(select(basic_stats_by_day,-(total_steps)))
tail(select(basic_stats_by_day,-(total_steps)))
```


## What is the average daily activity pattern?

```{r time_series, warning=FALSE}
# Group steps by 5-minute interval
mean_by_interval <- summarise(group_by(data, interval), mean_steps=mean(steps, na.rm = TRUE))

plot(mean_by_interval$interval, mean_by_interval$mean_steps, type = "l",
     xlab = "Time of day",
     ylab = "Mean number of steps",
     main = "Mean number of steps per five minute intervals in the day")
```

The time of day where the average number of steps is the maximum:

```{r max_day}
mean_by_interval$interval[which.max(mean_by_interval$mean_steps)]
```


## Imputing missing values

First, we need to know how many incomplete cases (rows with NAs) we have:

```{r incomplete cases}
sum(!complete.cases(data))
```

This number matches with the number of NAs in the *steps* column, displayed at the beginning of this document while summarising the data.

Given that it is only *steps* column that has missing values, I will use the mean number of steps by interval to replace the NAs.

```{r replacing NAs}
data_complete <- data
indices_na <- which(is.na(data_complete$steps))

for(i in indices_na) {
  inter <- data_complete$interval[i]
  mn <- mean_by_interval$mean_steps[mean_by_interval$interval == inter]
  data_complete[i,1] <- round(mn,digits = 0)
}
```

I will present a summary of the new data frame with the NAs replaced.

```{r summary_without_NAs}
# Get summary statistics by day
basic_stats_by_day_no_nas <- summarise(group_by(data_complete, date), total_steps=sum(steps), mean_steps=mean(steps), median_steps=median(steps))

# Draw the histogram
ggplot(basic_stats_by_day_no_nas, aes(x=date,y=total_steps)) + geom_bar(stat = 'identity') +
  ggtitle("Total number of steps per day (with replaced NAs)")
```

```{r print tables summary no-NAs, render=lemon_print}
# Print mean and median steps per day, head and tail to save space
head(select(basic_stats_by_day_no_nas,-(total_steps)))
tail(select(basic_stats_by_day_no_nas,-(total_steps)))
```

I will now calculate the impact on the total number of steps taken per day.

```{r impact}
basic_stats_by_day_diff <- basic_stats_by_day

# Repalce NAs with zeroes
basic_stats_by_day_diff[is.na(basic_stats_by_day_diff)] <- 0

# Finde difference between total number of steps per day
basic_stats_by_day_diff <- mutate(basic_stats_by_day_diff, diff = basic_stats_by_day_no_nas$total_steps - total_steps)

# Draw the histogram
ggplot(basic_stats_by_day_diff, aes(x=date,y=diff)) + geom_bar(stat = 'identity') + labs(x="Date",y="Difference steps") +
  ggtitle("Difference total number of steps per day (with replaced NAs)")

mean(basic_stats_by_day$total_steps, na.rm = TRUE)
mean(basic_stats_by_day_no_nas$total_steps, na.rm = TRUE)
```

The strategy used to replace NAs has a minimal impact on total steps per day and on the average of total steps per day. Only eight days are affected (days when there were no measurements) and this had a minimal effect on the average.


## Are there differences in activity patterns between weekdays and weekends?

First we need to group data by day of the week and compare the mean per day:

```{r weekdays}
# Create a new column with the day of the week
data_wd <- mutate(data, weekday = wday(date, week_start = 1))

# Get summary statistics by day of the week
basic_stats_by_wd <- summarise(group_by(data_wd, weekday), total_steps=sum(steps, na.rm = TRUE), mean_steps=mean(steps, na.rm = TRUE))
basic_stats_by_wd$weekday <- factor(basic_stats_by_wd$weekday)

days_of_week <- c("Mon","Tue","Wed","Thu","Fri","Sat","Sun")

# Draw the histogram
ggplot(basic_stats_by_wd, aes(x=weekday,y=mean_steps)) + geom_bar(stat = 'identity') + scale_x_discrete(labels= days_of_week) +
  labs(x="Day of the week",y="Total steps") +
  ggtitle("Total number of steps per day of the week")
```

It seems that the weekends are more active on average as expected. Now to analyse average daily activity by time of day and day of the week we can do this:

```{r weekday_activity}
# Group steps by 5-minute interval and day of the week
mean_by_interval_wd <- summarise(group_by(data_wd, weekday, interval), mean_steps=mean(steps, na.rm = TRUE))
mean_by_interval_wd$weekday <- factor(mean_by_interval_wd$weekday,       labels = days_of_week)

ggplot(mean_by_interval_wd, aes(x=interval,y=mean_steps)) + geom_line() +
  labs(x="Time of day",y="Mean steps") +
  facet_grid(weekday ~ .) +
  ggtitle("Mean number of steps per time of day and day of the week")
```

It seems that during the weekends the subject rises up later but is more consistently active during the entire day.